# Discrete Events Simulation with R

We are not going to explore this subject in depth, but at the end of this chapter we will understand what the discrete events simulation consists of, and we will carry out an example in R. Unlike Monte Carlo methods, which allow us to simulate events based on variables that change constantly over time, the discrete events simulation will help us to model the behavior of systems based on a sequence of discrete events over time. Remember that for both cases we use stochastic models.

Recall that different types of simulation apply depending on the nature of the system under consideration. A common model taxonomy classifies simulation problems along three main dimensions: 

1. deterministic vs. stochastic
2. static vs. dynamic (depending on whether they require a time component)
3. continuous vs. discrete (depending on how the system changes). 

For instance, Monte Carlo methods are well-known examples of static stochastic simulation techniques. On the other hand, discrete-event simulation (DES) is a specific technique for modelling stochastic, dynamic and discretely evolving systems. As opposed to continuous simulation, which typically uses smoothly-evolving equational models, DES is characterized by sudden state changes at precise points of (simulated) time.

Customers arriving at a bank, products being manipulated in a supply chain, or packets traversing a network are common examples of such systems. The discrete nature of a given system arises as soon as its behavior can be described in terms of events, which is the most fundamental concept in DES. An event is an instantaneous occurrence that may change the state of the system, while, between events, all the state variables remain constant.

There are several world views, or programming styles, for DES (Banks 2005): 

  - In the **activity-oriented** approach, a model consists of sequences of activities, or operations, waiting to be executed depending on some conditions. The simulation clock advances in fixed time increments. At each step, the whole list of activities is scanned, and their conditions, verified. Despite its simplicity, the simulation performance is too sensitive to the election of such a time increment.
  -Instead, the **event-oriented** approach completely bypasses this issue by maintaining a list of scheduled events ordered by time of occurrence. Then, the simulation just consists in jumping from event to event, sequentially executing the associated routines. 
  - Finally, the **process-oriented** approach refines the latter with the addition of interacting processes, whose activation is triggered by events. In this case, the modeller defines a set of processes, which correspond to entities or objects of the real system, and their life cycle.

In this course we will use the `simmer` package to perform our DES. `simmer` (Ucar and Smeets 2019a) is a DES package for R which enables high-level **process-oriented** modelling, in line with other modern simulators. But in addition, it exploits the novel concept of **trajectory**: a common path in the simulation model for entities of the same type. In other words, a trajectory consist of a list of standardized actions which defines the life cycle of equivalent processes. This design pattern is flexible and simple to use. It is important to note that `simmer` package uses pipes in order to simplify the structure of the simulation.

## `simmer` terminology

In this course we have already used some DES-specific terminology, e.g., event, state, entity, process or attribute. But there are, however, some simmer-specific terms, and some elements that require further explanation to understand the package architecture.

  + **Resource**: A passive entity, as it is commonly understood in standard DES terminology. However, simmer resources are conceived with queuing systems in mind, and therefore they comprise two internal self-managed parts:
    
    - Server: which, conceptually, represents the resource itself. It has a specified capacity and can be seized and released.
    - Queue: A priority queue of a certain size.

  + **Manager**: An active entity, i.e., a process, that has the ability to adjust properties of a resource (capacity and queue size) at run-time.
  + **Source**: A process responsible for creating new arrivals with a given inter-arrival time pattern and inserting them into the simulation model.
  + **Arrival**: A process capable of interacting with resources or other entities of the simulation model. It may have some attributes and prioritization values associated and, in general, a limited lifetime. Upon creation, every arrival is attached to a given trajectory.
  + **Trajectory**: An inter-linkage of activities constituting a recipe for arrivals attached to it, i.e., an ordered set of actions that must be executed. The simulation model is ultimately represented by a set of trajectories.
  + **Activity**: The individual unit of action that allows arrivals to interact with resources and other entities, perform custom routines while spending time in the system, move back and forth through the trajectory dynamically, and much more.

The `R` API (Application Programming Interface) exposed by `simmer` comprises two main elements: the **simmer environment** (or simulation environment) and the **trajectory object**.

## The Trajectory Object

A trajectory can be defined as a recipe and consists of an ordered set of activities. The `trajectory()` method instantiates the object, and activities can be appended using the `%>%` operator:

```{r}
library(simmer)
library(magrittr)

traj0 <- trajectory() %>%
  log_("Entering the trajectory") %>%
  timeout(10) %>%
  log_("Leaving the trajectory")
```

The trajectory above illustrates the two most basic activities available: displaying a message (`log_()`) and spending some time in the system (`timeout()`). 

An arrival attached to this trajectory will execute the activities in the given order, i.e., it will display “Entering the trajectory”, then it will spend 10 units of (simulated) time, and finally it will display “Leaving the trajectory”.

The example uses fixed parameters: a string and a numeric value respectively. However, at least the main parameter for all activities can also be what we will call a dynamical parameter, i.e., a function.

```{r}
traj1 <- trajectory() %>%
  log_(function() "Entering the trajectory") %>%
  timeout(10) %>%
  log_(function() "Leaving the trajectory")
```

Let's check what happens if we print the trajectory:

```{r}
traj1
```

We see the three activities we have created in our trajectory: message + delay + message. There are many activities available. We will briefly review them by categorizing them into different topics.

+ **Arrival properties**: Arrivals are able to store attributes and modify these using `set_attribute()`. Arrivals also hold a set of three prioritization values for accessing resources, `priority`, `preemptible` and `restart`, and we can use them inside the `set_priorization()` function.
+ **Interaction with resources**: The two main activities for interacting with resources are `seize()` and `release()`. In their most basic usage, they seize/release a given amount of a resource specified by name.
+ **Interaction with sources**: There are four activities specifically intended to modify arrival sources. An arrival may `activate()` or `deactivate()` a source, but also modify with `set_trajectory()` the trajectory to which it attaches the arrivals created, or set a new inter-arrival distribution with `set_source()`.
+ **Branching**: A branch is a point in a trajectory in which one or more sub-trajectories may be followed. Two types of branching are supported in `simmer.` The `branch()` activity places the arrival in one of the sub-trajectories depending on some condition evaluated in a dynamical parameter called option. On the other hand, the `clone()` activity is a parallel branch. It does not take any option.
+ **Loops**: There is a mechanism, `rollback()`, for going back in a trajectory and thus executing loops over a number of activities. This activity causes the arrival to step back a given amount of activities a number of times.
+ **Batching**: Batching consists of collecting a number of arrivals before they can continue their path in the trajectory as a unit 2 . This means that if, for instance, 10 arrivals in a batch try to seize a unit of a certain resource, only one unit may be seized, not 10.
+ **Asynchronous programming**: There are a number of methods enabling asynchronous events. The `send()` activity broadcasts one or more **signals** to all the arrivals subscribed to them. Signals can be triggered immediately or after some **delay**. Arrivals are able to block and `wait()` until a certain signal is received.
+ **Reneging**: Besides being rejected while trying to seize a resource, arrivals are also able to leave the trajectory at any moment, synchronously or asynchronously. Namely, reneging means that an arrival abandons the trajectory at a given moment. The most simple activity enabling this is `leave`, which immediately triggers the action given some probability.

## The Simulation Enviroment

The simulation environment manages resources and sources, and controls the simulation execution. The `simmer()` method instantiates the object, after which resources and sources can be appended using the `%>%` operator:

```{r}
env <- simmer() %>% 
  add_resource("res_name", 1) %>%
  add_generator("arrival", traj0, function() 25) %>%
  print()
```

Then, the simulation can be executed, or `run()`, until a stop time:

```{r}
env %>% run(until = 30)
```

There are a number of methods for extracting information, such as the simulation time (`now()`), future scheduled events (`peek()`), and getters for obtaining resources’ and sources’ parameters (capacity, queue size, server count and queue count; number of arrivals generated so far). There are also several setters available for resources and sources (capacity, queue size; trajectory, distribution).

+ **Resources**: A simmer resource comprises two internal self-managed parts: a **server** and a **priority queue**. Three main parameters define a resource: `name` of the resource, `capacity` of the server and `queue_size` (0 means no queue). Resources are monitored, non-preemptive and assumes a first-in-first-out (FIFO) policy by default.

+ **Sources**: Three main parameters define a source: a `name_prefix` for each generated arrival, a `trajectory` to attach them to and a source of inter-arrival times. There are two kinds of source: **generators** and **data sources**. A generator (`add_generator` method) is a dynamic source that draws inter-arrival times from a user-provided function. The `add_dataframe` method allows the user to set up a data source which draws arrivals from a provided data frame.

## Monitoring and data retrieval

There are three methods for obtaining monitored data about arrivals, resources and attributes. They can be applied to a single simulation environment or to a list of environments, and the returning object is always a data frame.

+ `get_mon_arrivals()`: Returns timing information per arrival: `name` of the arrival, `start_time`, `end_time`, `activity_time` (time not spent in resource queues) and a flag, `finished`, that indicates whether the arrival exhausted its activities (or was rejected). By default, this information is referred to the arrivals’ entire lifetime, but it may be obtained on a per-resource basis by specifying `per_resource = TRUE`.
+ `get_mon_resources()`: Returns state changes in resources.
+ `get_mon_attributes()`: Returns state changes in attributes.

## `simmer` Example

Let's see an example in R of discrete event simulation in which we simulate the behavior of patients in a hospital.

First we have to load the `simmer` and `magrittr` libraries to do the simulation

```{r}
library(simmer)
library(magrittr)
```

With the `simmer` library we create a simulation environment that we are going to call `outpatient clinic`

```{r}
env = simmer("outpatient clinic")
env
```

Second, we create the patient's trajectory. We do this using the `trajectory` function of the `simmer` library. 

Within this path we create the activities of the entities in the trajectory using the `seize` and `release` functions. Among these functions we introduce the time that each patient takes to carry out each event, that would correspond with the activity of the entities. 

For example, in the case of the nurse, we seize the entity with the function `seize`, we give it a time limit of 15 minutes (we do this simulating a random number with a normal distribution of mean 15 and standard deviation 1), and we release the entity with the function `release`.


```{r}
patient = trajectory("patients path") %>%
  seize("nurse", 1) %>%
  timeout(function() rnorm(1,15)) %>%
  release("nurse", 1) %>%
  
  seize("doctor", 1) %>%
  timeout(function() rnorm(1,20)) %>%
  release("doctor", 1) %>%
  
  seize("administration", 1) %>%
  timeout(function() rnorm(1,5)) %>%
  release("administration", 1)
```

Thirdly, we add the number of elements of the system using the `add_resource` function. 

In this case we introduce in the system two elements of the nurse type, three of the doctor type and two of the administration type. 

We also generate the patient entity that will interact with the whole system using the `add_generator` function. We will generate a patient every five minutes with a standard deviation of 0.5 minutes.


```{r}
env %>%
  add_resource("nurse", 2) %>%
  add_resource("doctor", 3) %>%
  add_resource("administration", 2) %>%
  add_generator("patient", patient, function() rnorm(1,5,0.5))
```

We finally launch the simulation for 540 minutes. When the simulation is finished we can see how our resources are working in the simulated system.

```{r}
env %>%
  run(until=540)
```

It is very easy to replicate a simulation multiple times using standard R functions.

```{r}
envs <- lapply(1:100, function(i) {
  simmer("outpatient clinic") %>%
    add_resource("nurse", 2) %>%
    add_resource("doctor", 3) %>%
    add_resource("administration", 2) %>%
    add_generator("patient", patient, function() rnorm(1,5,0.5)) %>%
    run(540)
})
```

This package provides some basic visualization tools to help you take a glance at your simulations quickly. There are three types of plot implemented with different metrics available:

  - Plot of resources. Two metrics available:
      * the usage of a resource over the simulation time frame.
      * the utilization of specified resources in the simulation.
  - Plot of arrivals. Three metrics available:
      * activity time.
      * waiting time.
      * flow time.
  - Plot of trayectories

With this graphs we can have a look at the overall resource utilization. The top and bottom of the error bars show respectively the 25th and 75th percentile of the utilization across all the replications. The top of the bar shows the median utilization.

```{r}
library(simmer.plot)

resources <- get_mon_resources(envs)
plot(resources, metric = "utilization")
```

It is also possible to have a look at resources’ activity during the simulation.

```{r}
plot(resources, metric = "usage", c("nurse", "doctor", "administration"), items = "server")
```

Next we can have a look at the evolution of the arrivals’ flow time during the simulation. In the plot below, each individual line represents a replication. A smooth line is drawn over them. All arrivals that didn’t finish their entire trajectory are excluded from the plot.

Similarly one can have a look at the evolution of the activity times with `metric = "activity_time"` and waiting times with `metric = "waiting_time"`.

```{r}
arrivals <- get_mon_arrivals(envs)
plot(arrivals, metric = "waiting_time")
```

Simulation trajectories may grow considerably, and they are not always easy to inspect to ensure their compliance with the model that we are trying to build. In this example we are going to expose a simple one:

```{r, echo=FALSE}
plot(patient)
```

Now that we know a bit about the `simmer` tool we are going to make some changes in the system to try to improve its efficiency. From what we have seen in the examples, it seems that the nurses are overworked. We will first try to "hire" a new nurse to improve the system.

```{r}
envs2 <- lapply(1:100, function(i) {
  simmer("outpatient clinic") %>%
    add_resource("nurse", 3) %>%
    add_resource("doctor", 3) %>%
    add_resource("administration", 2) %>%
    add_generator("patient", patient, function() rnorm(1,5,0.5)) %>%
    run(540)
})
```

Let's see how the system works with a new nurse through the plots:

```{r, echo=FALSE}
resources <- get_mon_resources(envs2)
plot(resources, metric = "utilization")
```

```{r, echo=FALSE}
plot(resources, metric = "usage", c("nurse", "doctor", "administration"), items = "server")
```

```{r, echo=FALSE}
arrivals <- get_mon_arrivals(envs2)
plot(arrivals, metric = "waiting_time")
```

We can see that both doctors and nurses are now working at 100% capacity. Now the problem seems to be coming from the administration. We are going to "fire" one of the workers in this area to try to improve the system.

```{r}
envs3 <- lapply(1:100, function(i) {
  simmer("outpatient clinic") %>%
    add_resource("nurse", 3) %>%
    add_resource("doctor", 3) %>%
    add_resource("administration", 1) %>%
    add_generator("patient", patient, function() rnorm(1,5,0.5)) %>%
    run(540)
})
```

Let's see how the system works without a administration worker through the plots:

```{r, echo=FALSE}
resources <- get_mon_resources(envs3)
plot(resources, metric = "utilization")
```

```{r, echo=FALSE}
plot(resources, metric = "usage", c("nurse", "doctor", "administration"), items = "server")
```

```{r, echo=FALSE}
arrivals <- get_mon_arrivals(envs3)
plot(arrivals, metric = "waiting_time")
```

Great! We've found the perfect balance for our little hospital.
